---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# predint

<!-- badges: start -->
<!-- badges: end -->

The package predint provides functions to calculate bootstrap calibrated 
prediction intervals for one or more future observations based on overdispersed 
binomial data, overdispersed poisson data, as well as data that is modeled by
linear random effects models fitted with lme4::lmer().

## Installation

You can install the released version of predint from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("predint")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("MaxMenssen/predint")
```
## Example
The following examples are based on the scenario described in Menssen and Schaarschmidt 
2019: Based on historical control data for the mortality of male B6C3F1-mice 
obtained in in long term studies at the National Toxicology Program, prediction 
intervals (PI) can be computed in order to verify the observed 
mortality of an actual (or future) trial. In this scenario prediction intervals 
can be comuted for two different puropses.  
On the one hand, a PI for one future observation can be computed in order to varify 
the outcome of one actual (or future) untreated control group that is compared
with several groups treated with the compund of interest.  
On the other hand, in some cases it might be usefull to varify the outcome of the complete
actual (or future) study based on the knowledge gained from historical control data.  
Simillarly to Menssen and Schaarschmidt 2019, it is assumed that the data is
overdispersed binomial. Hence, we will use the `quasi_bin_pi()` function in the following
two examples.

### Evaluation of one future control group
In this scenario, it is of interest to verify the control group of an actual (or future)
study that is comprised of 30 mice instead of 50 mice as in the historical data.
For this purpose a single prediction interval for one future observation is computed.

```{r example, cache=TRUE}
# load predint
library(predint)

# data set (Table 1 of the supplementary material of Menssen and Schaarschmidt 2019)
dat_real <- data.frame("dead"=c(15, 10, 12, 12, 13, 11, 19, 11, 14, 21),
                       "alive"=c(35, 40, 38, 38, 37, 39, 31, 39, 36, 29))

# PI for one future control group comprised of 50 mice
pi_m1 <- quasi_bin_pi(histdat=dat_real, newsize=30, traceplot = FALSE)
pi_m1
```

The historical binomial probability of success (historical mortality rate)
is `r pi_m1$hist_prob`, the bootstrap calibrated coefficient is `r pi_m1$quant_calib`
and the standard error of the prediction is `r pi_m1$pred_se`. The lower limit 
of the bootstrap calibrated asymptotic prediction interval is `r round(pi_m1$lower, 5)` 
and its upper limit is given by `r round(pi_m1$upper, 5)`.

If the mortality is  lower than `r round(pi_m1$lower, 5)` (practically spkoen lower than
`r ceiling(pi_m1$lower)`) it can be treated as unusual low. Consequently, mean 
comparisons between the control group might result in too many differences that 
are considered as significant and the compund of interest might be treated as more
hazardous than it actually is.  
On the other hand, the compund of interest might be treated as less hazardous 
if the mortality in the untreated control group is unusual high. This might be
the case, if its mortality exeeds`r round(pi_m1$upper, 5)` (practically spoken
higher than `r floor(pi_m1$upper)`).

### Evaluation of one future study
Let us assume, there is a study in which one untreated control group comprised of 50
male mice is compared to three treatment groups comprised of 30 mice each. If the
whole study should be compared to the historical knowledge, four prediction intervals
have to be computed. Hence `newsize` is set to `c(50, 30, 30, 30)`.

```{r, cache=TRUE}
pi_m4 <- quasi_bin_pi(histdat=dat_real, newsize=c(50, 30, 30, 30), traceplot = FALSE)
pi_m4
```

In this case, the untreated control group is in line with the historical control data
if its mortality falls between `r round(pi_m4$lower[1], 5)` and `r round(pi_m4$upper[1], 5)`. 
Simillarly, the groups treated with the compound of interest are in line with the historical knowlegde
regarding untreated control groups if their mortality ranges between `r round(pi_m4$lower[2], 5)`
and `r round(pi_m4$upper[2], 5)`. In this case inference based on such results might be highly 
questionable and more research about the compound of interest has to be done.

## References
Menssen M, Schaarschmidt F.: Prediction intervals for overdispersed binomial data
with application to historical controls. Statistics in Medicine. 2019;38:2652-2663.
https://doi.org/10.1002/sim.8124




